---
title: "smallRNA Biotype Comparison"
author: "Lance Parsons"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
editor: source
---

```{r}
#| label: load-packages
#| include: false
#| message: false
library("readr")
library("dplyr")
library("tidyr")
library("stringr")
library("ggplot2")
```

## Read data

Get list of samples

```{r}
samples <- read_tsv("config/samples.tsv",
  col_types = list(
    sample_name = col_character(),
    cell_line = col_factor(),
    exogenous_rna = col_factor(),
    day = col_factor()
  )
)
units <- read_tsv("config/units.tsv",
  col_types = list(
    sample_name = col_character(),
    unit_name = col_character(),
    fq1 = col_character(),
    fq2 = col_character()
  )
)
sample_units <- dplyr::left_join(samples, units, by = "sample_name") %>%
  unite(sample_unit, sample_name, unit_name, remove = FALSE) %>%
  mutate(
    day = as.factor(paste0("day", day)),
    batch = as.factor(case_when(
      exogenous_rna == "mastermix1" ~ "batch3",
      exogenous_rna == "mastermix2" ~ "batch3",
      TRUE ~ "batch2"
    )),
    .before = cell_line,
  )

sample_units
```

Table of exogenous RNA mixtures

```{r}
# Exogenous RNA mixtures
rna_mixes <- tibble()
for (mix in c("mastermix1", "mastermix2", "PJY103_mDNMT1", "PJY300_mDNMT1")) {
  t <- Biostrings::readDNAStringSet(sprintf("data/references/%s.fa", mix))
  rna_mixes <- rbind(rna_mixes, tibble(
    exogenous_rna = mix,
    rna_species = word(t@ranges@NAMES, 1),
    start = 1,
    end = t@ranges@width
  ))
}
rna_mixes
```

## Biotype comparison

* count only fragments that were properly aligned
* annotate with GENCODE gene model
* primary alignments were counted, even if the fragments aligned multiple times
* fragments aligning to multiple features were assigned to the feature that mostly closely overlapped with the fragment

```{r import_biotype_counts}
human_counts_dir <- "results/smrna_count/"
biotype_counts_files <- paste0(
  human_counts_dir,
  sample_units$sample_unit,
  "_first_proper_pair_biotype_count.txt"
)
exogenous_counts_dir <- "results/exogenous_rna_count/"
exogenous_counts_files <- paste0(
  exogenous_counts_dir,
  sample_units$sample_unit,
  "_idxstats.txt"
)

biotype_counts <- readr::read_tsv(
  biotype_counts_files[1],
  comment = "#",
  col_names = c("biotype", biotype_counts_files[1]),
  col_types = "ci"
)
exogenous_counts <- read_tsv(
  exogenous_counts_files[1],
  col_names = c("exogenous_rna", exogenous_counts_files[1]),
  col_types = "c-i-"
)
biotype_counts <- biotype_counts %>%
  dplyr::add_row(
    biotype = "exogenous_rna",
    "{biotype_counts_files[1]}" := # nolint: object_name_linter
      sum(exogenous_counts[[exogenous_counts_files[1]]])
  )

for (i in 2:length(biotype_counts_files)) {
  biotype_sample <-
    readr::read_tsv(
      biotype_counts_files[i],
      comment = "#",
      col_names = c("biotype", biotype_counts_files[i]),
      col_types = "ci"
    )

  exogenous_counts_sample <-
    readr::read_tsv(
      exogenous_counts_files[i],
      col_names = c("exogenous_rna", exogenous_counts_files[i]),
      col_types = "c-i-"
    )

  biotype_sample <- biotype_sample %>%
    dplyr::add_row(
      biotype = "exogenous_rna",
      "{biotype_counts_files[i]}" := # nolint: object_name_linter
        sum(exogenous_counts_sample[[exogenous_counts_files[i]]])
    )

  biotype_counts <- biotype_counts %>%
    dplyr::full_join(biotype_sample, by = "biotype")
}

biotype_counts <- biotype_counts %>%
  rename_all(~ stringr::str_replace_all(
    ., human_counts_dir, ""
  )) %>%
  rename_all(~ str_replace_all(
    .,
    "_first_proper_pair_biotype_count.txt",
    ""
  )) %>%
  tidyr::pivot_longer(!biotype, names_to = "sample", values_to = "count")

biotype_counts
```

```{r fig.height=7, fig.width=16}
#| fig-height: 7
#| fig-width: 16
p <- ggplot(
  data = subset(biotype_counts, !is.na(count)),
  aes(x = sample, y = count, fill = biotype)
) +
  geom_bar(stat = "identity", position = "fill") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p
```
