---
title: "smallRNA Biotype Comparison"
author: "Lance Parsons"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
editor: source
---


# Load libraries

This project uses [`renv`](https://rstudio.github.io/renv/articles/renv.html)
to keep track of installed packages. Install `renv` if not installed and load
dependencies with `renv::restore()`.

```r
install.packages("renv")
renv::restore()
```

```{r}
#| label: load-packages
#| include: false
#| message: false
library("readr")
library("dplyr")
library("tidyr")
library("stringr")
library("ggplot2")
```

# Read data

{{< include _sample-metadata.qmd >}}

# Biotype comparison

* count only fragments that were properly aligned
* annotate with GENCODE gene model
* primary alignments were counted, even if the fragments aligned multiple times
* fragments aligning to multiple features were assigned to the feature that mostly closely overlapped with the fragment

```{r}
#| label: import-biotype-counts
human_counts_dir <- "results/smrna_count/"
biotype_counts_files <- paste0(
  human_counts_dir,
  sample_units$sample_unit,
  "_first_proper_pair_biotype_count.txt"
)

biotype_counts <- readr::read_tsv(
  biotype_counts_files[1],
  comment = "#",
  col_names = c("biotype", biotype_counts_files[1]),
  col_types = "ci"
)

for (i in 2:length(biotype_counts_files)) {
  biotype_sample <-
    readr::read_tsv(
      biotype_counts_files[i],
      comment = "#",
      col_names = c("biotype", biotype_counts_files[i]),
      col_types = "ci"
    )

  biotype_counts <- biotype_counts %>%
    dplyr::full_join(biotype_sample, by = "biotype")
}

biotype_counts <- biotype_counts %>%
  rename_all(~ stringr::str_replace_all(
    ., human_counts_dir, ""
  )) %>%
  rename_all(~ str_replace_all(
    .,
    "_first_proper_pair_biotype_count.txt",
    ""
  )) %>%
  tidyr::pivot_longer(!biotype, names_to = "sample", values_to = "count")

biotype_counts
```

```{r fig.height=7, fig.width=16}
#| label: biotype-barplot
#| fig-height: 7
#| fig-width: 16
p <- ggplot(
  data = subset(biotype_counts, !is.na(count)),
  aes(x = sample, y = count, fill = biotype)
) +
  geom_bar(stat = "identity", position = "fill") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p
```
