---
title: "Adamson smallRNA - exogenous RNA"
output:
  html_notebook:
    toc: true
---

## Read data

This project uses [`renv`](https://rstudio.github.io/renv/articles/renv.html)
to keep track of installed packages. Install `renv` if not installed and load
dependencies with `renv::restore()`.

```r
install.packages("renv")
renv::restore()
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(pheatmap)
library(tibble)
library(ggplot2)
library(stringr)
library(cowplot)
library(markdown)
library(RColorBrewer)
```


Get list of samples

```{r}
samples <- str_remove(list.files("results/samtools_idxstats/"), ".bam.idxstats")
samples
```

Read idxstats to get grch38 coverage for normalization

```{r}
idxstats <- tibble(sample = samples, grch38_mapped_reads = NaN)
for (sample in samples) {

  # Read `idxstats` for normalization
  stats <- read_tsv(sprintf(
    "results/samtools_idxstats/%s.bam.idxstats",
    sample
  ),
  col_names = c(
    "sequence_name", "sequence_length",
    "mapped_reads", "unmapped_reads"
  ),
  col_types = "ciii"
  )
  grch38_mapped_reads <- stats %>%
    filter(!sequence_name %in% c("PJY142", "PJY151", "PJY209", "*")) %>%
    select(mapped_reads) %>%
    sum()
  idxstats <- rows_update(idxstats,
    tibble(
      sample = sample,
      grch38_mapped_reads = grch38_mapped_reads
    ),
    by = "sample"
  )
}
idxstats
```

Read `bedpe` files to get exogenous target coverage

```{r}
bedpe_data <- tibble()
for (sample in samples) {
  data <-
    readr::read_tsv(sprintf(
      "results/exogenous_targets_bedpe/%s_exogenous_targets.bedpe", sample
    ),
    col_names = c(
      "chrom1", "chrom1Start", "chrom1End",
      "chrom2", "chrom2Start", "chrom2End",
      "name", "score", "strand1", "strand2"
    ),
    col_types = "ciiciicicc"
    )
  bedpe_data <- tibble(rbind(
    bedpe_data,
    cbind(
      filter(idxstats, sample == !!sample),
      data
    )
  ))
}
bedpe_data
```

Summarize data

* Count only pairs that start at position 0
* Sum for each "end" position
* Normalize using reads mapped to human genome (grch38)

```{r}
summarized_data <- bedpe_data %>%
  filter(chrom1Start == 0) %>%
  count(sample, grch38_mapped_reads, chrom1, chrom1Start, chrom2End,
    name = "count"
  ) %>%
  mutate(normalizedCount = (count * 10^6) / grch38_mapped_reads)
summarized_data
```



Reorganize data for correlation analysis

```{r}
corr_data <- summarized_data %>%
  select(-grch38_mapped_reads, -count) %>%
  pivot_wider(names_from = sample, values_from = normalizedCount) %>%
  arrange(chrom1, chrom1Start, chrom2End)
corr_data
```

Save summarized data tables to text files

```{r}
write_tsv(summarized_data, file = "exogenous_targets_paired_read_summary.tsv")
write_tsv(corr_data, file = "exogenous_targets_normalized_counts.tsv")
```


## Replicate correlation

```{r fig.height=6, fig.width=6}
p <- ggplot(
  corr_data,
  aes(
    x = PJY142_PJY99_rep2,
    y = PJY142_PJY99_rep3,
  )
) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "gray") +
  ggtitle("PJY142_PJY99")
p
```

```{r fig.height=6, fig.width=6}
p <- ggplot(
  corr_data,
  aes(
    x = PJY142_PJY209_rep2,
    y = PJY142_PJY209_rep3,
  )
) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "gray") +
  ggtitle("PJY142_PJY209")
p
```

```{r fig.height=6, fig.width=6}
p <- ggplot(
  corr_data,
  aes(
    x = PJY151_PJY99_rep2,
    y = PJY151_PJY99_rep3,
  )
) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "gray") +
  ggtitle("PJY151_PJY99")
p
```

```{r fig.height=6, fig.width=6}
p <- ggplot(
  corr_data,
  aes(
    x = PJY151_PJY209_rep2,
    y = PJY151_PJY209_rep3,
  )
) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "gray") +
  ggtitle("PJY151_PJY209")
p
```


---

## Exogenous species comparison

### PJY142

Histogram of fragments starting at position zero (0)
```{r, fig.width=12}
pjy142_length <- 113
pjy142_hist <-
  ggplot(
    summarized_data %>%
      filter(chrom1 == "PJY142") %>%
      complete(
        nesting(sample, grch38_mapped_reads, chrom1, chrom1Start),
        chrom2End = seq(0, pjy142_length),
        fill = list(count = 0, normalizedCount = 0)
      ),
    aes(x = chrom2End, y = normalizedCount, fill = sample)
  ) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  facet_grid(vars(sample))
plot_grid(pjy142_hist, pjy142_hist + scale_y_continuous(trans = "log1p"))
```

Heatmap colors
```{r}
blue_colors <- colorRampPalette(brewer.pal(9, "Blues"))(100)
```

Heatmap PJY142
```{r, fig.width=6, fig.height=14}
pheatmap(corr_data %>%
  filter(chrom1 == "PJY142") %>%
  complete(
    nesting(chrom1, chrom1Start),
    chrom2End = seq(0, pjy142_length),
    fill = list(count = 0, normalizedCount = 0)
  ) %>%
  unite(seqid, chrom1, chrom1Start, chrom2End) %>%
  column_to_rownames("seqid") %>%
  log1p(),
cluster_rows = FALSE,
color = blue_colors,
fontsize_row = 6,
)
```


### PJY151

Histogram of fragments starting at position zero (0)
```{r, fig.width=12}
pjy151_length <- 113
pjy151_hist <-
  ggplot(
    summarized_data %>%
      filter(chrom1 == "PJY151") %>%
      complete(
        nesting(sample, grch38_mapped_reads, chrom1, chrom1Start),
        chrom2End = seq(0, pjy151_length),
        fill = list(count = 0, normalizedCount = 0)
      ),
    aes(x = chrom2End, y = normalizedCount, fill = sample)
  ) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  facet_grid(vars(sample))
plot_grid(pjy151_hist, pjy151_hist + scale_y_continuous(trans = "log1p"))
```

Heatmap PJY151
```{r, fig.width=6, fig.height=15}
pheatmap(corr_data %>%
  filter(chrom1 == "PJY151") %>%
  complete(
    nesting(chrom1, chrom1Start),
    chrom2End = seq(0, pjy151_length),
    fill = list(count = 0, normalizedCount = 0)
  ) %>%
  unite(seqid, chrom1, chrom1Start, chrom2End) %>%
  column_to_rownames("seqid") %>%
  log1p(),
cluster_rows = FALSE,
color = blue_colors,
fontsize_row = 6
)
```

### PJY99

Histogram of fragments starting at position zero (0)
```{r, fig.width=12}
pjy99_length <- 138
pjy99_hist <-
  ggplot(
    summarized_data %>%
      filter(chrom1 == "PJY209", chrom1Start == 0) %>%
      filter(grepl("PJY99", sample)) %>%
      filter(chrom2End <= pjy99_length) %>%
      complete(
        nesting(sample, grch38_mapped_reads, chrom1, chrom1Start),
        chrom2End = seq(0, pjy99_length),
        fill = list(count = 0, normalizedCount = 0)
      ),
    aes(x = chrom2End, y = normalizedCount, fill = sample)
  ) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  facet_grid(vars(sample))
plot_grid(pjy99_hist, pjy99_hist + scale_y_continuous(trans = "log1p"))
```


Heatmap PJY99
```{r, fig.width=6, fig.height=20}
pheatmap(corr_data %>%
  filter(chrom1 == "PJY209", chrom1Start == 0) %>%
  select(!contains("PJY209")) %>%
  filter(chrom2End <= pjy99_length) %>%
  complete(
    nesting(chrom1, chrom1Start),
    chrom2End = seq(0, pjy99_length),
    fill = list(count = 0, normalizedCount = 0)
  ) %>%
  unite(seqid, chrom1, chrom1Start, chrom2End) %>%
  column_to_rownames("seqid") %>%
  log1p(),
cluster_rows = FALSE,
color = blue_colors,
)
```



### PJY209

Histogram of fragments starting at position zero (0)
```{r, fig.width=12}
pjy209_length <- 174
pjy209_hist <-
  ggplot(
    summarized_data %>%
      filter(chrom1 == "PJY209", chrom1Start == 0) %>%
      filter(grepl("PJY209", sample)) %>%
      complete(
        nesting(sample, grch38_mapped_reads, chrom1, chrom1Start),
        chrom2End = seq(0, pjy209_length),
        fill = list(count = 0, normalizedCount = 0)
      ),
    aes(x = chrom2End, y = normalizedCount, fill = sample)
  ) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  facet_grid(vars(sample))
plot_grid(pjy209_hist, pjy209_hist + scale_y_continuous(trans = "log1p"))
```

Heatmap PJY209
```{r, fig.width=6, fig.height=20}
pheatmap(corr_data %>%
  filter(chrom1 == "PJY209") %>%
  select(!contains("PJY99")) %>%
  complete(
    nesting(chrom1, chrom1Start),
    chrom2End = seq(0, pjy209_length),
    fill = list(count = 0, normalizedCount = 0)
  ) %>%
  unite(seqid, chrom1, chrom1Start, chrom2End) %>%
  column_to_rownames("seqid") %>%
  log1p(),
cluster_rows = FALSE,
color = blue_colors,
fontsize_row = 6
)
```
