---
title: "Adamson smallRNA - exogenous RNA"
output:
  html_notebook:
    toc: true
---

## Read data

This project uses [`renv`](https://rstudio.github.io/renv/articles/renv.html)
to keep track of installed packages. Install `renv` if not installed and load
dependencies with `renv::restore()`.

```r
install.packages("renv")
renv::restore()
```

```{r message=FALSE}
library(markdown)
library(stringr)
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
```


Get list of samples

```{r}
samples <- str_remove(list.files("results/samtools_idxstats/"), ".bam.idxstats")
samples
```

Read idxstats to get grch38 coverage for normalization

```{r}
idxstats <- tibble(sample = samples, grch38_mapped_reads = NaN)
for (sample in samples) {

  # Read `idxstats` for normalization
  stats <- read_tsv(sprintf(
    "results/samtools_idxstats/%s.bam.idxstats",
    sample
  ),
  col_names = c(
    "sequence_name", "sequence_length",
    "mapped_reads", "unmapped_reads"
  ),
  col_types = "ciii"
  )
  grch38_mapped_reads <- stats %>%
    filter(!sequence_name %in% c("PJY142", "PJY151", "PJY209", "*")) %>%
    select(mapped_reads) %>%
    sum()
  idxstats <- rows_update(idxstats,
    tibble(
      sample = sample,
      grch38_mapped_reads = grch38_mapped_reads
    ),
    by = "sample"
  )
}
idxstats
```

Read gene biotype summary stats

```{r}
gene_biotype_summary <- tibble(gene_biotype = character())
for (sample in samples) {
  stats <- read_tsv(sprintf(
    "results/homo_sapiens_gene_biotype_summary/%s.gene_biotype.summary",
    sample
  ),
  col_names = c(
    "gene_biotype", sample
  ),
  col_types = "ci"
  )
  gene_biotype_summary <-
    gene_biotype_summary %>% full_join(stats, by = "gene_biotype")
}
gene_biotype_summary
```


Tidy data for easier manipulation and plotting

```{r}
gene_biotype_summary_tidy <- gene_biotype_summary %>%
  pivot_longer(-gene_biotype) %>%
  rename(sample = name, count = value) %>%
  group_by(sample) %>%
  mutate(freq_in_sample = round(count / sum(count), 3)) %>%
  arrange(sample, desc(count))
gene_biotype_summary_tidy
```


Plot any biotype with at least 1% frequency in a sample

```{r}
ggplot(
  gene_biotype_summary_tidy %>%
    group_by(gene_biotype) %>%
    filter(any(freq_in_sample > 0.01)),
  aes(
    x = reorder(gene_biotype, -count),
    y = freq_in_sample,
    fill = sample
  )
) +
  geom_bar(stat = "identity", position = position_dodge()) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```

Plot any biotype with at least 1% frequency in a sample

```{r}
ggplot(
  gene_biotype_summary_tidy %>%
    mutate(gene_biotype = reorder(
      gene_biotype, count, sum,
      decreasing = FALSE
    )) %>%
    group_by(gene_biotype) %>%
    filter(any(freq_in_sample > 0.01)) %>%
    droplevels(),
  aes(
    x = sample,
    y = count,
    fill = gene_biotype
  )
) +
  geom_bar(stat = "identity", position = "fill") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```

Save summarized data tables to text files

```{r}
write_tsv(gene_biotype_summary,
  file = "all_sample_gene_biotype_summary.tsv"
)
write_tsv(gene_biotype_summary_tidy,
  file = "all_sample_gene_biotype_summary_tidy.tsv"
)
```
