---
title: "Quantification of Three Prime Coverage"
author: "Lance Parsons"
date: last-modified
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
    fig-format: svg
editor: source
---

# Load libraries

This project uses `renv` to keep track of installed packages. Install `renv` if not installed and load dependencies with `renv::restore()`.

``` r
install.packages("renv")
renv::restore()
```

```{r}
#| label: load-packages
#| include: false
#| message: false
library("readr")
library("dplyr")
library("Rsamtools")
library("tidyr")
library("stringr")
library("forcats")
library("GenomicAlignments")
library("plyranges")
library("ggplot2")
```

# Read sample data

{{< include _sample-metadata.qmd >}}

# Exogenous RNA counts

{{< include _exogenous-alignments.qmd >}}

# Human small RNA gene counts

{{< include _human-small-rna-counts.qmd >}}

## Human small RNA gene totals

```{r}
#| label: sum-human-small-rna
human_rna_totals <- human_gene_counts %>%
  pivot_longer(
    cols = -"gene",
    names_to = "sample_unit",
    values_to = "count"
  ) %>%
  group_by(.data$sample_unit) %>%
  summarize(count = sum(.data$count, na.rm = TRUE))

### Export human RNA totals
write_tsv(human_rna_totals, "human_rna_count_totals.tsv")
human_rna_totals
```


# 3' Coverage Quantification

Here we attempt to quantify the difference in the cis-active coverage on the three-prime (3') between the parental and SSB-ko4 samples. This is done by first counting the number of cis-active fragments for each sample that contain the first nucleotide of the primer binding site (PBS) and the number of cis-active fragments that do not cover that base. Then, for each sample we normalize these counts to the total human smRNA and finally compute the fraction of cis-active fragments containing the first nucleotide of the PBS.

```{r}
#| label: cis-actitve-pbs-quantification

# Create table with: sample, active_cis_pbs, active_cis_nopbs
three_prime_quanitification <- tibble(
  sample_unit = character(),
  sequence_name = character(),
  cis_active_pbs = integer(),
  cis_active_nopbs = integer(),
  cis_active_full_sgrna = integer(),
  cis_active_partial_sgrna = integer(),
  human_small_rna = integer(),
)
for (rna_species in names(rna_species_plot_data)) {
  rna_species_data <- rna_species_plot_data[[rna_species]]
  for (sample_unit in names(rna_species_data)) {
    sample_unit_data <- rna_species_data[[sample_unit]]

    # Read annotations for PBS location
    annotations <- read_tsv(
      sprintf("data/references/exogenous-rna/%s.bed", rna_species),
      col_names = c("chrom", "chromStart", "chromEnd", "name"),
      col_types = "ciic"
    )
    pbs_start <- annotations %>%
      filter(name == "PBS") %>%
      pull(chromStart)
    sgrna_end <- annotations %>%
      filter(name == "sgRNA_end") %>%
      pull(chromStart)
    cis_active_pbs <- length(sample_unit_data$active_cis %>% plyranges::filter(end > pbs_start))
    cis_active_nopbs <- length(sample_unit_data$active_cis %>% plyranges::filter(end <= pbs_start))
    cis_active_full_sgrna <- length(sample_unit_data$active_cis %>% plyranges::filter(end >= sgrna_end))
    cis_active_partial_sgrna <- length(sample_unit_data$active_cis %>% plyranges::filter(end < sgrna_end))
    human_small_rna <- normalization_factor(sample_unit, rna_species, "active_cis", "human_small_rna")
    three_prime_quanitification <- three_prime_quanitification %>%
      add_row(
        sample_unit = sample_unit,
        sequence_name = rna_species,
        cis_active_pbs = cis_active_pbs,
        cis_active_nopbs = cis_active_nopbs,
        cis_active_full_sgrna = cis_active_full_sgrna,
        cis_active_partial_sgrna = cis_active_partial_sgrna,
        human_small_rna = human_small_rna,
      )
  }
}
three_prime_quanitification <- three_prime_quanitification %>%
  mutate(norm_cis_active_pbs = cis_active_pbs / human_small_rna) %>%
  mutate(norm_cis_active_nopbs = cis_active_nopbs / human_small_rna) %>%
  mutate(fraction_cis_active_pbs = cis_active_pbs / (cis_active_pbs + cis_active_nopbs)) %>%
  mutate(norm_cis_active_full_sgrna  = cis_active_full_sgrna  / human_small_rna) %>%
  mutate(norm_cis_active_partial_sgrna = cis_active_partial_sgrna / human_small_rna) %>%
  mutate(fraction_cis_active_sgrna = cis_active_full_sgrna / (cis_active_full_sgrna + cis_active_partial_sgrna)) %>%
  full_join(sample_units, by = join_by(sample_unit))
three_prime_quanitification
write_tsv(three_prime_quanitification, "three-prime-quantification.tsv")
```


```{r}
#| label: normalize-cis-active-pbs

three_prime_quanitification_summarized <-
  three_prime_quanitification %>%
  dplyr::group_by(sequence_name, cell_line, day) %>%
  summarize(
    cis_active_pbs_mean = mean(cis_active_pbs),
    cis_active_pbs_se = sd(cis_active_pbs) / sqrt(length(cis_active_pbs)),
    norm_cis_active_pbs_mean = mean(norm_cis_active_pbs),
    norm_cis_active_pbs_se = sd(norm_cis_active_pbs) / sqrt(length(norm_cis_active_pbs)),
    cis_active_nopbs_mean = mean(cis_active_nopbs),
    cis_active_nopbs_se = sd(cis_active_nopbs) / sqrt(length(cis_active_nopbs)),
    norm_cis_active_nopbs_mean = mean(norm_cis_active_nopbs),
    fraction_cis_active_pbs_mean = mean(fraction_cis_active_pbs),
    fraction_cis_active_pbs_se = sd(fraction_cis_active_pbs) / sqrt(length(fraction_cis_active_pbs)),
    #
    cis_active_full_sgrna_mean = mean(cis_active_full_sgrna),
    cis_active_full_sgrna_se = sd(cis_active_full_sgrna) / sqrt(length(cis_active_full_sgrna)),
    norm_cis_active_full_sgrna_mean = mean(norm_cis_active_full_sgrna),
    norm_cis_active_full_sgrna_se = sd(norm_cis_active_full_sgrna) / sqrt(length(norm_cis_active_full_sgrna)),
    cis_active_partial_sgrna_mean = mean(cis_active_partial_sgrna),
    cis_active_partial_sgrna_se = sd(cis_active_partial_sgrna) / sqrt(length(cis_active_partial_sgrna)),
    norm_cis_active_partial_sgrna_mean = mean(norm_cis_active_partial_sgrna),
    fraction_cis_active_sgrna_mean = mean(fraction_cis_active_sgrna),
    fraction_cis_active_sgrna_se = sd(fraction_cis_active_sgrna) / sqrt(length(fraction_cis_active_sgrna)),
    .groups = "keep"
  )
three_prime_quanitification_summarized
write_tsv(three_prime_quanitification_summarized, "three-prime-quantification-summarized.tsv")

```


```{r}
#| label: plot_threeprime-quantification-pbs

p <- ggplot(
  summarized_three_prime_quantification %>%
    mutate(sample_day = interaction(sequence_name, day)),
  aes(
    fill = cell_line,
    y = fraction_cis_active_pbs_mean,
    x = sample_day,
  )
) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_errorbar(aes(
    ymin = fraction_cis_active_pbs_mean - fraction_cis_active_pbs_se,
    ymax = fraction_cis_active_pbs_mean + fraction_cis_active_pbs_se
  ), width = .2, position = position_dodge(.9)) +
  ggtitle("PBS Coverage") +
  ylab("Fraction of reads covering PBS start") +
  xlab("Sample/Day") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0))
p
```


```{r}
#| label: plot_threeprime-quantification-sgrna

p <- ggplot(
  summarized_three_prime_quantification %>%
    mutate(sample_day = interaction(sequence_name, day)),
  aes(
    fill = cell_line,
    y = fraction_cis_active_sgrna_mean,
    x = sample_day,
  )
) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_errorbar(aes(
    ymin = fraction_cis_active_sgrna_mean - fraction_cis_active_sgrna_se,
    ymax = fraction_cis_active_sgrna_mean + fraction_cis_active_sgrna_se
  ), width = .2, position = position_dodge(.9)) +
  ggtitle("sgRNA Coverage") +
  ylab("Fraction of reads covering sgRNA end") +
  xlab("Sample/Day") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0))
p
```

```{r}
#| label: three-prime-counts-function

cis_active_threeprime_counts <- function(rna_species_plot_data, annotation, with_operator, without_operator, offset = 0) {
  with_fun = match.fun(with_operator)
  without_fun = match.fun(without_operator)
  with_annotation_name <- sprintf("cis_active_with_%s", annotation)
  without_annotation_name <- sprintf("cis_active_without_%s", annotation)
  if (offset != 0) {
    with_annotation_name <- sprintf("%s_%s", with_annotation_name, offset)
    without_annotation_name <- sprintf("%s_%s", without_annotation_name, offset)
  }
  # Create table with: sample, with_annotation, without_annotation, human_small_rna
  three_prime_quanitification <- tibble(
    sample_unit = character(),
    sequence_name = character(),
    "{with_annotation_name}" := integer(),
    "{without_annotation_name}" := integer(),
    human_small_rna = integer(),
  )
  for (rna_species in names(rna_species_plot_data)) {
    rna_species_data <- rna_species_plot_data[[rna_species]]
    for (sample_unit in names(rna_species_data)) {
      sample_unit_data <- rna_species_data[[sample_unit]]
  
      # Read annotations
      annotations <- read_tsv(
        sprintf("data/references/exogenous-rna/%s.bed", rna_species),
        col_names = c("chrom", "chromStart", "chromEnd", "name"),
        col_types = "ciic"
      )
      annotation_location <- annotations %>%
        filter(name == annotation) %>%
        pull(chromStart) + offset
      cis_active_with_annotation <- length(sample_unit_data$active_cis %>% 
                                             plyranges::filter(with_fun(end, annotation_location)))
      cis_active_without_annotation <- length(sample_unit_data$active_cis %>% 
                                             plyranges::filter(without_fun(end, annotation_location)))
      human_small_rna <- normalization_factor(sample_unit, rna_species, "active_cis", "human_small_rna")
      three_prime_quanitification <- three_prime_quanitification %>%
        add_row(
          sample_unit = sample_unit,
          sequence_name = rna_species,
          "{with_annotation_name}" := cis_active_with_annotation,
          "{without_annotation_name}" := cis_active_without_annotation,
          human_small_rna = human_small_rna,
        )
    }
  }
  three_prime_quanitification <- three_prime_quanitification %>%
    mutate("norm_{with_annotation_name}" := .data[[with_annotation_name]]  / human_small_rna) %>%
    mutate("norm_{without_annotation_name}" := .data[[without_annotation_name]]  / human_small_rna) %>%
    mutate("fraction_{with_annotation_name}" := .data[[with_annotation_name]] / (.data[[with_annotation_name]] + .data[[without_annotation_name]])) %>%
    full_join(sample_units, by = join_by(sample_unit))
  return(three_prime_quanitification)
}

# t <- cis_active_threeprime_counts(rna_species_plot_data, "PBS", ">", "<=")
```

```{r}
#| label: summarize-three-prime-counts-function

summarize_three_prime_counts <- function(three_prime_quanitification, with_column, without_column) {
  summarized_data <- three_prime_quanitification %>%
    dplyr::group_by(sequence_name, cell_line, day) %>%
    summarize(
      "{with_column}_mean" := mean( .data[[with_column]] ),
      "{with_column}_se" := sd( .data[[with_column]] ) / sqrt(length(.data[[with_column]])),
      
      "norm_{with_column}_mean" := mean( .data[[paste0("norm_", with_column)]] ),
      "norm_{with_column}_se" := sd( .data[[paste0("norm_", with_column)]] ) / sqrt(length(.data[[paste0("norm_", with_column)]])),
      
      "{without_column}_mean" := mean( .data[[without_column]] ),
      "{without_column}_se" := sd( .data[[without_column]] ) / sqrt(length(.data[[without_column]])),
      
      "norm_{without_column}_mean" := mean( .data[[paste0("norm_", without_column)]] ),
      "norm_{without_column}_se" := sd( .data[[paste0("norm_", without_column)]] ) / sqrt(length(.data[[paste0("norm_", without_column)]])),
      
      "fraction_{with_column}_mean" := mean(.data[[paste0("fraction_", with_column)]]),
      "fraction_{with_column}_se" := sd(.data[[paste0("fraction_", with_column)]]) / sqrt(length( .data[[paste0("fraction_", with_column)]] )),
      .groups = "keep"
  )
  return(summarized_data)
}

# ts <- summarize_three_prime_counts(t, "cis_active_with_PBS", "cis_active_without_PBS")
# ts
# write_tsv(three_prime_quanitification_summarized, "three-prime-quantification-summarized.tsv")
```



```{r}
#| label: three-prime-counts-pbs
threeprime_dir = "threeprime-counts"
dir.create(threeprime_dir, showWarnings = FALSE)
threeprime_pbs_counts <- cis_active_threeprime_counts(rna_species_plot_data, "PBS", ">", "<=")
write_tsv(threeprime_pbs_counts, file.path(threeprime_dir, "threeprime-pbs-counts.tsv"))
threeprime_pbs_counts
```

```{r}
#| label: summarize-three-prime-counts-pbs
threeprime_pbs_counts_summarized <- summarize_three_prime_counts(
  threeprime_pbs_counts, "cis_active_with_PBS", "cis_active_without_PBS")
write_tsv(threeprime_pbs_counts_summarized, file.path(threeprime_dir, "threeprime-pbs-counts-summarized.tsv"))
threeprime_pbs_counts_summarized
```


```{r}
#| label: plot_threeprime-count-pbs

p <- ggplot(
  threeprime_pbs_counts_summarized %>%
    mutate(sample_day = interaction(sequence_name, day)),
  aes(
    fill = cell_line,
    y = fraction_cis_active_with_PBS_mean,
    x = sample_day,
  )
) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_errorbar(aes(
    ymin = fraction_cis_active_with_PBS_mean - fraction_cis_active_with_PBS_se,
    ymax = fraction_cis_active_with_PBS_mean + fraction_cis_active_with_PBS_se
  ), width = .2, position = position_dodge(.9)) +
  ggtitle("PBS Coverage") +
  ylab("Fraction of reads covering PBS start") +
  xlab("Sample/Day") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0))
p
```



```{r}
#| label: three-prime-counts-sgrna

threeprime_sgrna_counts <- list()
for (offset in 0:-5) {
  offset_label <- paste0("offset", offset)
  threeprime_dir = "threeprime-counts"
  dir.create(threeprime_dir, showWarnings = FALSE)
  threeprime_sgrna_counts[[offset_label]] <- 
    cis_active_threeprime_counts(rna_species_plot_data, "sgRNA_end", ">", "<=", offset = offset)
  write_tsv(threeprime_sgrna_counts[[offset_label]],
            file.path(threeprime_dir, sprintf("threeprime-sgrna-counts-%s.tsv", offset_label)))
  print(threeprime_sgrna_counts[[offset_label]])
}
```

```{r}
#| label: summarize-three-prime-counts-pbs

threeprime_sgrna_counts_summarized <- list()
for (offset in 0:-5) {
  offset_label <- paste0("offset", offset)
  offset_suffix <- paste0("_", offset)
  if (offset == 0) {offset_suffix <- ""}

  threeprime_sgrna_counts_summarized[[offset_label]] <- summarize_three_prime_counts(
    threeprime_sgrna_counts[[offset_label]], 
    paste0("cis_active_with_sgRNA_end",offset_suffix),
    paste0("cis_active_without_sgRNA_end", offset_suffix))
  write_tsv(threeprime_sgrna_counts_summarized[[offset_label]],
            file.path(threeprime_dir, sprintf("threeprime-sgrna-counts-%s-summarized.tsv", offset_label)))
  print(threeprime_sgrna_counts_summarized[[offset_label]])
}
```

```{r}
#| label: plot_threeprime-count-sgrna
for (offset in 0:-5) {
  offset_label <- paste0("offset", offset)
  offset_suffix <- paste0("_", offset)
  if (offset == 0) {offset_suffix <- ""}
  
  mean_col <- paste0("fraction_cis_active_with_sgRNA_end", offset_suffix, "_mean")
  se_col <- paste0("fraction_cis_active_with_sgRNA_end", offset_suffix, "_se")
  p <- ggplot(
    threeprime_sgrna_counts_summarized[[offset_label]] %>%
      mutate(sample_day = interaction(sequence_name, day)),
    aes(
      fill = cell_line,
      y = .data[[mean_col]],
      x = sample_day,
    )
  ) +
    geom_bar(position = "dodge", stat = "identity") +
    geom_errorbar(aes(
      ymin = .data[[mean_col]] - .data[[se_col]],
      ymax = .data[[mean_col]] + .data[[se_col]]
    ), width = .2, position = position_dodge(.9)) +
    ggtitle("sgRNA Coverage") +
    ylab(sprintf("Fraction of reads covering sgRNA end (%s)", offset)) +
    xlab("Sample/Day") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0))
  print(p)
}
```
